\textit{[něco jako?]} Generating lambda term individuals using undisciplined A*

### 2013-10-22 tex/evogp.tex
evogp tex

\newcommand{\Lets}{Let us\xspace}
\newcommand{\lets}{let us\xspace}
\newcommand{\lterm}{$\lambda$-term\xspace}
\newcommand{\lterms}{$\lambda$-terms\xspace}
\newcommand{\lhead}{$\lambda$-head\xspace}
\newcommand{\lheads}{$\lambda$-heads\xspace}
\newcommand{\la}{\leftarrow\xspace}
\newcommand{\Lp}  {\Lambda^{\prime}\xspace}
\newcommand{\tur}[3]{#1\vdash{}#2:#3}
\newcommand{\turst}[3]{$#1\vdash{}#2:#3$\xspace}
\newcommand{\GMS}{\turst{\Gamma}{M}{\sigma}}
\newcommand{\atTree}{@-tree\xspace}
\newcommand{\setDots}[2]{ \lbrace #1 , \dots , #2 \rbrace}
\newcommand{\lh}[1]{\lambda #1}
\newcommand{\sexprTree}{sexpr-tree\xspace}
\newcommand{\SexprTree}{Sexpr-tree\xspace}
\newcommand{\then}{\Rightarrow\xspace}
\newcommand{\lamb}[2]{( \lambda \, #1 \, . \, #2 )}
\newcommand{\lam}[2]{\lambda \, #1 \, . \, #2}
\newcommand{\ST}{\mathop{\mathrm{ST}}}
\newcommand{\FV}{\mathop{\mathrm{FV}}}
\newcommand{\Scomb }{\mathbf{S}}
\newcommand{\Kcomb }{\mathbf{K}}
\newcommand{\Icomb }{\mathbf{I}}
\newcommand{\bbarr}{\twoheadrightarrow_\beta}
\newcommand{\barr}{\rightarrow_\beta}
\newcommand{\beq}{=_\beta}
\newcommand{\eearr}{\twoheadrightarrow_\eta}
\newcommand{\earr}{\rightarrow_\eta}
\newcommand{\eeq}{=_\eta}
\newcommand{\bearr}{\rightarrow_{\beta\eta}}
\newcommand{\bbeearr}{\twoheadrightarrow_{\beta\eta}}
\newcommand{\beeq}{=_{\beta\eta}}
\newcommand{\etar}{\twoheadrightarrow_\eta}
\newcommand{\ered}{$\eta$-reduction\xspace}
\newcommand{\bnf}{$\beta$-\textit{nf}\xspace}
\newcommand{\enf}{$\eta$-\textit{nf}\xspace}
\newcommand{\eenf}{$\eta^{-1}$-\textit{nf}\xspace}
\newcommand{\beenf}{$\beta\eta^{-1}$-\textit{nf}\xspace}
\newcommand{\benf}{$\beta\eta$-\textit{nf}\xspace}
\newcommand{\bredex}{$\beta$-redex\xspace} 
\newcommand{\lnf}{\textit{lnf}\xspace}
\newcommand{\Ae}{\mathop{\mathrm{\AE}}}
\newcommand{\Bcomb }{\mathbf{B}}   
\newcommand{\BBcomb }{\mathbf{B*}}
\newcommand{\Ccomb }{\mathbf{C}}   
\newcommand{\CCcomb }{\mathbf{C'}}
\newcommand{\SScomb }{\mathbf{S'}}
\newcommand{\ar}{\rightarrow\xspace}
\newcommand{\T}{\mathbb{T}\xspace}
\newcommand{\Real}{\mathbb{R}}

@@

\title{\textit{[něco jako?]} 
Generating lambda term individuals using undisciplined A*}
\author{Tom\'{a}\v{s} K\v{r}en \and Roman Neruda}
\institute{Matfyz}
\maketitle

\begin{abstract}
\textit{(Aby tu něco bylo, pak se udělá lepší (trochu upravenej abstract z diplomky))}
In this paper, generalization of the standard genetic programming (GP)
for simply typed lambda calculus is presented. We use population 
initialization method parameterized by simple search strategy. 
First described strategy corresponds to standard ramped half-and-half method, 
second one corresponds to exhaustive systematic search and third one is a 
novel geometric strategy, which outperforms standard method in success 
rate, time consumption and average individual size in two experiments. 
Other performance enhancements based on theory of lambda calculus are 
proposed and supported by experiment. Abstraction elimination is 
utilized to enable use of simple tree-swapping crossover.
\end{abstract}

@@

@@

@@

@@

@@

@@

\textit{- Unifished term je term který může obsahovat jako list uzel tvaru 
$(\tau,\Gamma)$, takzvaný unfinished leaf (UL).\\}

@@

The inputs for the term generating algorithm are following.
\begin{enumerate}
 \item Desired type $\tau$ of generated terms.
 \item Context $\Gamma$ representing set of building symbols.
 \item Number $n$ of terms to be generated.
\end{enumerate}

\textit{
- Pracujeme s prioritní frontou rozdělaných lambda termů.\\ 
- Priorita je dána počtem UL v termu.\\
- Na počátku je do fronty vložen jediný UT = $(\tau,\Gamma)$\\
- Z fronty vyndám term s nejnižším \#UL, pokud má \#UL = 0, pak je to vygenerovaný jedinec, pokud ne tak vezmu jeho DFS-první UL a expanduju ho. Výsledkem expandování je několik (possibly 0) unfin. termů, které jsou až na expandovaný uzel totožné s původním UT, daný UT je nahrazen novým pod-UT. \\
- Expandování probíhá následovně: \\
(a) je to  $(\sigma \ar \tau,\Gamma)$ pak tento list nahradim novým podstromem $\lam{x}{(\tau;\Gamma,x:a)}$\\
(b) je to $(\alpha,\Gamma)$ pak pro každý\\ 
$f : (\tau_1 \ar \dots \ar \tau_n \ar \alpha) \in \Gamma$ \\
nahradíme tento list novým podstromem \\
$(~f~(\tau_1,\Gamma)~\dots~(\tau_n,\Gamma)~)$
\\\\
- Prohledávací strategie má pak funkci filtru následníků. \\
- Pokud se fronta vyprázdní ještě před vygenerováním požadovaného 
  počtu termů tak zbytek dogenerujeme stejným zpusobem, tzn do 
  fronty přihodíme $(\tau,\Gamma)$ a jedem dál \\
- Systematická strategie nezahodí nic (Pro ní generování 
  kolapsuje do A* algoritmu).\\
- Ramped-half-and-half zahodí vždy vše krom jediného.\\
- Geometrická strategie nezahodí s pravděpodobností q na hloubka UL \\
}

\textbf{$\eta$-normalization ...}

@@

\subsection{Crossover???}

@@

Desired type of generated programs $\sigma$ and 
building blocks context $\Gamma$ are following.
\begin{align*}
\sigma = \Real \ar &\Real\\
\Gamma = \{
  (+)  &: \Real \ar \Real \ar \Real    ,
  (-)   : \Real \ar \Real \ar \Real    ,
  (*)   : \Real \ar \Real \ar \Real    ,
  rdiv  : \Real \ar \Real \ar \Real    ,\\
  sin  &: \Real \ar \Real              ,
  cos   : \Real \ar \Real              ,
  exp   : \Real \ar \Real              , 
  rlog  : \Real \ar \Real              \}
\end{align*}
where
\begin{align*}
rdiv(p,q) &= \begin{cases} 1 &\mbox{if } q = 0 \\
p/q & \mbox{otherwise } \end{cases}  \\
rlog(x) &= \begin{cases} 0 &\mbox{if } x = 0 \\
log(\vert x\vert) & \mbox{otherwise}. \end{cases}
\end{align*}

Fitness function is computed as follows

>>>>

\newcommand{\Lets}{Let us\xspace}
\newcommand{\lets}{let us\xspace}
\newcommand{\lterm}{$\lambda$-term\xspace}
\newcommand{\lterms}{$\lambda$-terms\xspace}
\newcommand{\lhead}{$\lambda$-head\xspace}
\newcommand{\lheads}{$\lambda$-heads\xspace}
\newcommand{\la}{\leftarrow\xspace}
\newcommand{\Lp}  {\Lambda^{\prime}\xspace}
\newcommand{\tur}[3]{#1\vdash{}#2 \colon #3}
\newcommand{\turst}[3]{$#1\vdash{}#2:#3$\xspace}
\newcommand{\GMS}{\turst{\Gamma}{M}{\sigma}}
\newcommand{\atTree}{@-tree\xspace}
\newcommand{\setDots}[2]{ \lbrace #1 , \dots , #2 \rbrace}
\newcommand{\lh}[1]{\lambda #1}
\newcommand{\sexprTree}{sexpr-tree\xspace}
\newcommand{\SexprTree}{Sexpr-tree\xspace}
\newcommand{\then}{\Rightarrow\xspace}
\newcommand{\lamb}[2]{( \lambda \, #1 \, . \, #2 )}
\newcommand{\lam}[2]{\lambda \, #1 \, . \, #2}
\newcommand{\ST}{\mathop{\mathrm{ST}}}
\newcommand{\FV}{\mathop{\mathrm{FV}}}
\newcommand{\Scomb }{\mathbf{S}}
\newcommand{\Kcomb }{\mathbf{K}}
\newcommand{\Icomb }{\mathbf{I}}
\newcommand{\bbarr}{\twoheadrightarrow_\beta}
\newcommand{\barr}{\rightarrow_\beta}
\newcommand{\beq}{=_\beta}
\newcommand{\eearr}{\twoheadrightarrow_\eta}
\newcommand{\earr}{\rightarrow_\eta}
\newcommand{\eeq}{=_\eta}
\newcommand{\bearr}{\rightarrow_{\beta\eta}}
\newcommand{\bbeearr}{\twoheadrightarrow_{\beta\eta}}
\newcommand{\beeq}{=_{\beta\eta}}
\newcommand{\etar}{\twoheadrightarrow_\eta}
\newcommand{\ered}{$\eta$-reduction\xspace}
\newcommand{\bnf}{$\beta$-\textit{nf}\xspace}
\newcommand{\enf}{$\eta$-\textit{nf}\xspace}
\newcommand{\eenf}{$\eta^{-1}$-\textit{nf}\xspace}
\newcommand{\beenf}{$\beta\eta^{-1}$-\textit{nf}\xspace}
\newcommand{\benf}{$\beta\eta$-\textit{nf}\xspace}
\newcommand{\bredex}{$\beta$-redex\xspace} 
\newcommand{\lnf}{\textit{lnf}\xspace}
\newcommand{\Ae}{\mathop{\mathrm{\AE}}}
\newcommand{\Bcomb }{\mathbf{B}}   
\newcommand{\BBcomb }{\mathbf{B*}}
\newcommand{\Ccomb }{\mathbf{C}}   
\newcommand{\CCcomb }{\mathbf{C'}}
\newcommand{\SScomb }{\mathbf{S'}}
\newcommand{\ar}{\rightarrow\xspace}
\newcommand{\T}{\mathbb{T}\xspace}
\newcommand{\Real}{\mathbb{R}}

\newcommand{\ul}[2]{\langle #1 ; #2 \rangle}

@@

\title{\textit{[něco jako?]} 
Generating lambda term individuals using undisciplined A*}
\author{Tom\'{a}\v{s} K\v{r}en \and Roman Neruda}
\institute{Matfyz ...}
\maketitle

\begin{abstract}
\textbf{\textit{
(Aby tu něco bylo, pak se udělá lepší 
(tohle je trochu upravenej abstract z diplomky))}}
In this paper, generalization of the standard genetic programming (GP)
for simply typed lambda calculus is presented. We use population 
initialization method parameterized by simple search strategy. 
First described strategy corresponds to standard ramped half-and-half method, 
second one corresponds to exhaustive systematic search and third one is a 
novel geometric strategy, which outperforms standard method in success 
rate, time consumption and average individual size in two experiments. 
Other performance enhancements based on theory of lambda calculus are 
proposed and supported by experiment. Abstraction elimination is 
utilized to enable use of simple tree-swapping crossover.
\end{abstract}

@@

(...)\\

@@

~\\(...)

@@

\subsection{Ramped half-and-half}

@@

\subsubsection{Reductions}
\subsubsection{Tree representations}
%\subsubsection{}

@@

\textit{--- nezapomenout popsat $\Gamma,x:\tau$ notaci
 a to že to automaticky implikuje že x neni obsažena v $\Gamma$,
 tzn je nová...}

\subsection{\GMS}

@@

\textit{- Unifished term je term který může obsahovat jako list uzel tvaru 
$(\tau,\Gamma)$, takzvaný unfinished leaf (UL).\\
-Vysvětlit unfinished leaf.\\}

@@

The inputs for the term generating algorithm are following.
\begin{enumerate}
 \item Desired type $\tau$ of generated terms.
 \item Context $\Gamma$ representing set of building symbols.
 \item Number $n$ of terms to be generated.
 \item Search strategy $S$. 
\end{enumerate}

Essential data structure of our term generating algorithm 
is priority queue of unfinished terms. 
Priority of an unfinished term is given by 
the number of unfinished leafs in that term.
Initially, the queue contains only one unfinished term; 
$\ul{\tau}{\Gamma}$.\\

At each step, the term $M$ with the lowest number of 
unfinished leafs is pulled from the queue.
According to the actual number of those leafs one of
the following actions is performed.
\begin{enumerate}
 \item If the term has no unfinished leaf (i.e., it is a finished
 term satisfying \mbox{$\tur{\Gamma}{M}{\tau}$}), then it is added to the
 result collection of generated terms.   
 \item Otherwise, \textit{successors} of the unfinished term $M$ are
       filtered out by \textit{search strategy} $S$ and
       those successors that outlast the filtration 
       are inserted into the queue.
\end{enumerate}

\textit{Successors} of an unfinished term $M$ are obtained by 
\textit{expansion} of the \mbox{\textit{DFS-first}} unfinished leaf $L$
(i.e., the leftmost unfinished leaf of $M$).\\

Expansion of the selected unfinished leaf $L$ leads to creation of 
one or many (possibly zero) successors.
In this process, $L$ is replaced
by a new subterm defined by the following rules.
\begin{enumerate}
 \item If $L = \ul{\rho \ar \sigma}{\Gamma}$, 
       then $L$ is replaced by $\lamb{x}{\ul{\sigma}{\Gamma,x \colon \rho}}$.
       Thus this expansion results in exactly one successor.  
 \item If $L = \ul{\alpha}{\Gamma}$ where $\alpha$ is \textit{atomic} type,
       then for each 
       \mbox{$f : (\tau_1 \ar \dots \ar \tau_n \ar \alpha) \in \Gamma$}
       the unfinished leaf $L$ is replaced by 
       $(~f~(\tau_1,\Gamma)~\dots~(\tau_n,\Gamma)~)$.
       Thus this expansion results in many (possibly zero or one) successors.
\end{enumerate}

Now that we have all possible successors of $M$, we are about to apply
the \textit{search strategy} $S$. A search strategy is a procedure
which takes as input a set of unfinished terms and returns a subset
of the input set. Therefore, search strategy acts as a filter reducing 
the search space.

If the queue becomes empty before the desired number $n$ of terms
is generated, then the initial unfinished term $\ul{\tau}{\Gamma}$ 
is inserted to the queue and the process continues.

\Lets now discuss three such search strategies.

\subsubsection{Systematic strategy}
\subsubsection{Ramped half-and-half strategy}
\subsubsection{Geometric strategy}

~\\\\
\textit{
- Systematická strategie nezahodí nic (Pro ní generování 
  kolapsuje do A* algoritmu).\\
- Ramped-half-and-half zahodí vždy vše krom jediného.\\
- Geometrická strategie nezahodí s pravděpodobností q na hloubka UL \\
}

\subsection{Discussion and further improvements}

\subsection{$\eta$-normalization}

@@

\subsection{Crossover}

\textit{Zmínil bych jí tu ale jen hodně rychle...}

@@

Desired type of generated programs $\sigma$ and 
building blocks context $\Gamma$ are following.
\begin{align*}
\sigma = \Real \ar &\Real\\
\Gamma = \{
  (+)  &: \Real \ar \Real \ar \Real    ,
  (-)   : \Real \ar \Real \ar \Real    ,
  (*)   : \Real \ar \Real \ar \Real    ,
  rdiv  : \Real \ar \Real \ar \Real    ,\\
  sin  &: \Real \ar \Real              ,
  cos   : \Real \ar \Real              ,
  exp   : \Real \ar \Real              , 
  rlog  : \Real \ar \Real              \}
\end{align*}
where

\noindent\begin{minipage}{.5\linewidth}
\begin{align*}
rdiv(p,q) &= \begin{cases} 1 &\mbox{if } q = 0 \\
p/q & \mbox{otherwise } \end{cases}  
\end{align*}
\end{minipage}%
\begin{minipage}{.5\linewidth}
\begin{align*}
rlog(x) &= \begin{cases} 0 &\mbox{if } x = 0 \\
log(\vert x\vert) & \mbox{otherwise}. \end{cases}
\end{align*}
\end{minipage}

~\\\\
Fitness function is computed as follows
